---
title: "ACCESSIBILITY ANALYSIS: Data preps"
author: "JMG"
date: "2024-03-18"
output: html_document
---

# DATA PREPARATION

This involves, importing, re-projection and clipping of datasets: Admin boundaries, road network, barriers [lakes and rivers], land cover, DEM, population density and healthcare facilities

## libraries

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(sf)
library(osmextract)
library(tmap)
library(rgdal)
library(dplyr)
library(terra)
library(doParallel)
library(foreach)
library(snow)
library(raster)

```

# Uganda boundary

Used for clipping land cover raster

```{r}
admin0 <- st_read("D:/PHP_spring_2024/SSCI 680/final project/datasets/UG admin boundaries/uga_admbnda_adm0_ubos_20200824.shp")
st_crs(admin0)

#admin_proj <- st_crs(admin)$proj4string
admin0 <- admin0 %>% st_transform("+proj=utm +zone=36 +south +datum=WGS84 +units=m +no_defs +type=crs")


admin_proj <- st_crs(admin0)$proj4string # save the projection

# visualize Uganda boundary
tm_shape(admin0)+
  tm_polygons()

#st_write(admin0, "admin0.shp")
```

## Road network data from OSM

Only primary, secondary and tertiary roads were used while modeling travel time in AccessMod

```{r}
## using ?oe_get function in osmextract package
road_osm = oe_get("uganda",
             query = "SELECT * FROM 'lines' WHERE highway IS NOT NULL",
             quiet=F, force_download=T)

class(road_osm)
View(road_osm)

road_osm <- road_osm %>% dplyr::select(-c(4:8)) # remove unwanted colunms


### some reclassification
road_osm <- road_osm %>% mutate(class_n = highway)
class(road_osm$class_n)
road_osm$class_n <- as.factor(road_osm$class_n)
road_osm$class_n[road_osm$class_n=="motorway_link"]<-"motorway"
road_osm$class_n[road_osm$class_n=="primary_link"]<-"primary"
road_osm$class_n[road_osm$class_n=="secondary_link"]<-"secondary"
road_osm$class_n[road_osm$class_n=="trunk_link"]<-"trunk"
road_osm$class_n[road_osm$class_n=="tertiary_link"]<-"tertiary"

## remove some road classifications: irrelevant to the study area
road_osm <- road_osm %>% filter(class_n != "bridleway")
road_osm <- road_osm %>% filter(class_n != "construction")
road_osm <- road_osm %>% filter(class_n != "passing_place")
road_osm <- road_osm %>% filter(class_n != "proposed")
road_osm <- road_osm %>% filter(class_n != "rest_area")
road_osm <- road_osm %>% filter(class_n != "bridleway")
road_osm <- road_osm %>% filter(class_n != "road")
road_osm <- road_osm %>% filter(class_n != "services")
road_osm <- road_osm %>% filter(class_n != "steps")
road_osm <- road_osm %>% filter(class_n != "yes")

#droplevels
road_osm$class_n <-droplevels(road_osm$class_n)

## add another column for labels
road_osm <- road_osm %>% mutate(labels = class_n)

road_osm$labels <- factor(road_osm$labels,
        levels = c("motorway","trunk","primary","secondary",
                   "tertiary","unclassified","residential",
                   "footway","living_street","track","pedestrian",
                   "service","path","cycleway"),
        labels = c(1,2,3,4,5,6,7,8,9,10,11,12,13,14))

#my road network
roadnetwork <- road_osm %>% dplyr::select(c(1,6,7,5))

## rename some columns
roadnetwork <-   roadnetwork %>% 
  rename(
    label = class_n,
    class = class_n2
    )
## coercise class to numeric
roadnetwork$class <- as.integer(roadnetwork$class)

## reproject roadnetwork data
roadnetwork <- st_transform(roadnetwork, crs("+proj=utm +zone=36 +south +datum=WGS84 +units=m +no_defs"))

### check road types
### in the analysis i intend to use primary, secondary, and tertiary, roads only. These road classification are more accurate 

table(roadnetwork$labels)

## export this road network
st_write(roadnetwork, "roadnetwork.shp")



## map
tm_shape(admin0) +
  tm_polygons() +
   tm_shape(roadnetwork) +
  tm_lines ()

View(roadnetwork)

head(roadnetwork)
```

## Land cover data

Raster at 100m resolution, 2019 Downloaded from Copernicus Global Land Service

```{r}
# load landcover data
landcover <- rast("D:/PHP_spring_2024/SSCI 680/final project/datasets/landcover/PROBAV_LC100_global_v3.0.1_2019-nrt_Discrete-Classification-map_EPSG-4326.tif")
class(landcover)

res(landcover) # check resolution

# clip raster land cover layer to Uganda admin boundary
UGlandcover <- crop(landcover, extent(admin0))
UGlandcover <- mask(UGlandcover, admin0)
plot(UGlandcover)


UGlandcover <- project(UGlandcover, "+proj=utm +zone=36 +south +datum=WGS84 +units=m +no_defs", method="near") #re-projection

plot(UGlandcover)


### save raster land cover file at tif
writeRaster(UGlandcover, filename = "UGlandcover.tif")


## map landcover
tm_shape(admin0) +
  tm_polygons() +
   tm_shape(UGlandcover) +
  tm_raster ()

```

## Barriers to movement: rivers and lakes

This data could be downloaded from OSM. However i used shape files already available from ICPAC Geoportal

```{r}

### river data from icpac geoportal
rivers <- st_read("D:/PHP_spring_2024/SSCI 680/final project/datasets/rivers/3a37081d-1f4e-434e-b056-ca5e86ed7f62/uga_water_lines_dcw.shp")
st_crs(rivers)$proj4string

#View(rivers)

### re-projection
rivers_utm <- st_transform(rivers, crs("+proj=utm +zone=36 +south +datum=WGS84 +units=m +no_defs"))

tm_shape(admin2) +
  tm_polygons() +
   tm_shape(rivers_utm) +
  tm_lines ()

##### save rivers as shape file
st_write(rivers_utm, "rivers.shp")
```

## Lake data from ICPAC geoportal

```{r}
############################### lakes from ICPAC geoportal
lakes <- st_read("D:/PHP_spring_2024/SSCI 680/final project/datasets/lakes/4b767c70-97e4-461d-a691-0e13fbf359bf/uga_water_areas_dcw.shp")

st_crs(lakes)$proj4string
#View(lakes)

### re-projection
lakes_utm <- st_transform(lakes, crs("+proj=utm +zone=36 +south +datum=WGS84 +units=m +no_defs"))

tm_shape(admin2) +
  tm_polygons() +
   tm_shape(lakes_utm) +
  tm_polygons ()

##### save lakes as shape file
st_write(lakes_utm, "lakes.shp")

```

## DEM

obtained from the Africa geoportal. Prepared by The Shuttle Radar Topography Mission data at 30m resolution for Uganda. This raster layer was resampled to a 100m resolution. 100m is the resolution i used to model accessibility in AccessMod.

```{r}
srtm <- rast("D:/PHP_spring_2024/SSCI 680/final project/datasets/DEM/Uganda_SRTM30meters.tif")
st_crs(srtm)$proj4string
plot(srtm)

##re-projection
srtm_UTM <- srtm %>% terra::project("+proj=utm +zone=36 +south +datum=WGS84 +units=m +no_defs", method="near")
plot(srtm_UTM)

## change DEM resolution from 30 m to 100m
# Resample the raster to match the resolution of the reference raster (landcover raster)
srtm_UTM <- resample(srtm_UTM, UGlandcover, method = "bilinear")

## save dem as USGSDEM file
terra::writeRaster(srtm_UTM, filename = "UGDEM2.img", overwrite = TRUE)

```

## Population data

The raster data was produced by WorldPop Research Group at the University of Southampton This data is at a 1km resolution. i resampled it to 100m resolution.

```{r}
pop_dat <- rast("D:/PHP_spring_2024/SSCI 680/final project/datasets/uga_pd_2020_1km.tif")
plot(pop_dat)
st_crs(admin)$proj4string

## re-projection pop 
pop_dat <- pop_dat %>% project("+proj=utm +zone=36 +south +datum=WGS84 +units=m +no_defs", method="near")  


## save pop raster
writeRaster(pop_dat, filename = "pop_rast_1km.tif")


# Resample the raster to match the resolution of the reference raster (landcover raster)
pop_dat_100m <- resample(pop_dat, UGlandcover, method = "bilinear")
plot(pop_dat_100m)

## save pop raster 100m
writeRaster(pop_dat_100m, filename = "pop_rast_100m.tif")

```

## District admin shapefile

this data contains district boundaries its relevant when conducting zonal analysis in AccessMod. zonal analysis is used i estimate the proportion of the population able to access a healthcare facility within a given travel time.

```{r}
ug_admin2 <- st_read("D:/PHP_spring_2024/SSCI 680/final project/datasets/UG admin boundaries/uga_admbnda_adm2_ubos_20200824.shp")

## reproject admin2 data
ug_admin2 <- st_transform(ug_admin2, crs("+proj=utm +zone=36 +south +datum=WGS84 +units=m +no_defs"))
st_crs(ug_admin2)$proj4string

tm_shape(admin2)+
  tm_polygons()+
  tm_shape(ug_admin2)+
  tm_polygons()

#save admin2 as shp
st_write(ug_admin2, "UGadmin2.shp")
```

## Health facility data

This data contains health facility data of different types. It was obtained from the ministry of health Uganda.

```{r}
facilities <- st_read("D:/PHP_spring_2024/SSCI 680/final project/datasets/healthhfacilities/healthcenters.shp")

#View(facilities)

####re-projection
facilities_utm <- st_transform(facilities, crs("+proj=utm +zone=36 +south +datum=WGS84 +units=m +no_defs"))
st_crs(facilities_utm)$proj4string

# facilities per district
tm_shape(admin2)+
  tm_polygons()+
  tm_shape(facilities_utm)+
  tm_dots ()

#save facilities as shp
st_write(facilities_utm, "facilities.shp")
```

# ANALYSIS OF ACCESSIBILITY FOR SELECTED REGIONS

Uganda has 4 major regions, but for my analysis, i focused on 2 regions (central and eastern). I subsetted the landcover, roadnetwork, facilities, and population datasets at a regional level. After data preparation, this was imported into AccessMod

## UG regions

```{r}
admin1 <- st_read("D:/PHP_spring_2024/SSCI 680/final project/datasets/UG admin boundaries/uga_admbnda_adm1_ubos_20200824.shp")

admin1 <- admin1 %>% st_transform("+proj=utm +zone=36 +south +datum=WGS84 +units=m +no_defs +type=crs")

# visualize the regions
tm_shape(admin1)+
  tm_polygons()

file_name <- admin1$ADM1_EN# to integrate country name in saving step later

## boundaries by regions, save them as shape files for used later
system.time(for(i in 1:nrow(admin1)) {
  a = admin1 %>% 
    filter(ADM1_EN == file_name[i])
  a= st_as_sf(a)
  st_write(a, paste0("Reg_", file_name[i], ".shp"), overwrite=T)
}) 

# visualize central region
Reg_C <- st_read("D:/PHP_spring_2024/SSCI 680/final project/SSCI680 final project data prep/Reg_Central.shp")

tm_shape(Reg_C)+
  tm_polygons()

```

## Creating land cover for each region and saving the raster layers

```{r}
#import UG landcover rater
UGlandcover <- rast("D:/PHP_spring_2024/SSCI 680/final project/SSCI680 final project data prep/UGlandcover.tif")

file_name <- admin1$ADM1_EN# to integrate country name in saving step later

system.time(for(i in 1:nrow(admin1)) {
  l = UGlandcover
  a = admin1 %>% 
    filter(ADM1_EN == file_name[i])
  a = vect(a)
  r = terra::crop(l, a)
  m = terra::mask(r, a)
  st_write(m, paste0("Reglandcover_", file_name[i], ".tif"), overwrite=T)
}) 


# visualize landcover in Central region
Reglandcover_central <- rast("D:/PHP_spring_2024/SSCI 680/final project/SSCI680 final project data prep/Reglandcover_Central.tif")

tm_shape(Reglandcover_central) +
  tm_raster ()

plot(Reglandcover_central)

```

## Creating DEM for each region and saving the raster layers

```{r}
## UGDEM layer
UGDEM2 <- rast("D:/PHP_spring_2024/SSCI 680/final project/SSCI680 final project data prep/UGDEM2.img")

file_name <- admin1$ADM1_EN# to integrate country name in saving step later

system.time(for(i in 1:nrow(admin1)) {
  l = UGDEM2
  a = admin1 %>% 
    filter(ADM1_EN == file_name[i])
  a = vect(a)
  r = terra::crop(l, a)
  m = terra::mask(r, a)
  writeRaster(m, paste0("RegDEM_", file_name[i], ".tif"), overwrite=T)
}) 


# visualize DEM in Central region
RegDEM_central <- rast("D:/PHP_spring_2024/SSCI 680/final project/SSCI680 final project data prep/RegDEM_Central.tif")

tm_shape(RegDEM_central) +
  tm_raster ()


```

## Create population raster for each region

```{r}
## pop raster layer
regpop100m <- rast("D:/PHP_spring_2024/SSCI 680/final project/SSCI680 final project data prep/pop_rast_100m.tif")

file_name <- admin1$ADM1_EN# to integrate country name in saving step later

system.time(for(i in 1:nrow(admin1)) {
  l = regpop100m
  a = admin1 %>% 
    filter(ADM1_EN == file_name[i])
  a = vect(a)
  r = terra::crop(l, a)
  m = terra::mask(r, a)
  writeRaster(m, paste0("RegPop_", file_name[i], ".tif"), overwrite=T)
}) 


# visualize population in Central region
RegPop_central <- rast("D:/PHP_spring_2024/SSCI 680/final project/SSCI680 final project data prep/RegPop_Central.tif")

tm_shape(Reglandcover_central) +
  tm_raster ()

```

## Create road network for each region

```{r}
# import road network
roadnetwork <- st_read("D:/PHP_spring_2024/SSCI 680/final project/SSCI680 final project data prep/roadnetwork.shp")

file_name <- admin1$ADM1_EN# to integrate country name in saving step later

system.time(for(i in 1:nrow(admin1)) {
  l = roadnetwork
  a = admin1 %>% 
    filter(ADM1_EN == file_name[i])
  #a = vect(a)
  a= st_as_sf(a)
  r = sf::st_intersection(l, a)
  #m = terra::mask(r, a)
  #p = terra::project(m, admin1$prj4, method = "near")
  #values(p)[values(p) == 0] <- NA
  st_write(r, paste0("Regroadnetwork_", file_name[i], ".shp"), overwrite=T)
}) 

# visualize roadnetwork in central region
Regroadnetwork_C <- st_read("D:/PHP_spring_2024/SSCI 680/final project/SSCI680 final project data prep/Regroadnetwork_Central.shp")

tm_shape(Regroadnetwork_C)+
  tm_lines()


```

## Create healthcare facilities for each region

```{r}
# import facilities
UGfacilities <- st_read("D:/PHP_spring_2024/SSCI 680/final project/SSCI680 final project data prep/facilities.shp")

file_name <- admin1$ADM1_EN# to integrate country name in saving step later

system.time(for(i in 1:nrow(admin1)) {
  l = UGfacilities
  a = admin1 %>% 
    filter(ADM1_EN == file_name[i])
  a= st_as_sf(a)
  r = sf::st_intersection(l, a)
  st_write(r, paste0("Regfacilities_", file_name[i], ".shp"), overwrite=T)
}) 


# visualize facilities in central region
Regfacilities_C <- st_read("D:/PHP_spring_2024/SSCI 680/final project/SSCI680 final project data prep/Regfacilities_Central.shp")
tm_shape(Reg_C)+
  tm_polygons()+
tm_shape(Regfacilities_C)+
  tm_dots()
```
